A Plasma Torus Around a Young Low-Mass Star

### 2025-04-03 paper/ms.tex
update

\documentclass{nature3}
\usepackage{graphicx}
\usepackage{float}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{aas_macros_nature}
\usepackage{lineno}
\usepackage{caption}

@@

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

% * keck/hires: behavior at v/veq<1
The data in the H$\alpha$ line core are more complex.  At $|\Delta v /
v_{\rm eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material may additionally modulate the
line profile.
Figure~\ref{fig:spec}e suggests line core variability caused by both
bright and dark regions on the star's surface, superposed
with smaller-amplitude variability from the transiting circumstellar
material.
For instance, from $\phi$=0-0.3, the double-peaked emission feature
is visible when it is being viewed both on and off-limb.
However, the large emission feature that crosses the star from
$\phi$=0.4-0.9 cannot be circumstellar in origin; it emits at an
amplitude greater than that observed from the circumstellar
components, and it crosses the stellar velocity surface too slowly.
It must therefore originate from a chromospherically bright region
near the stellar surface.
While the origin of the other bright and dark streaks passing across the
line core are more ambiguous,
a final exercise to quantify the behavior of this region is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

@@

@@

>>>>

\documentclass{nature3}
\usepackage{graphicx}
\usepackage{float}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{aas_macros_nature}
\usepackage{lineno}
\usepackage{caption}
\usepackage{rotating} % sidewaystable

@@

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of this optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

% * keck/hires: behavior at v/veq<1
The H$\alpha$ line core is more complex.  At $|\Delta v /
v_{\rm eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material may additionally modulate the
line profile.
Figure~\ref{fig:spec}e suggests line core variability caused by both
bright and dark regions on the star's surface, superposed
with smaller-amplitude variability from the transiting circumstellar
material.
For instance, from $\phi$=0-0.3, the double-peaked emission feature
is visible when it is being viewed both on and off-limb.
However, the large emission feature that crosses the star from
$\phi$=0.4-0.9 cannot be circumstellar in origin; it emits at an
amplitude greater than that observed from the circumstellar
components, and it crosses the stellar velocity surface too slowly.
It must therefore originate from a chromospherically bright region
near the stellar surface.
While the origin of the other bright and dark streaks passing across the
line core are more ambiguous,
a final exercise to quantify the behavior of this region is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

@@

\subsection{Detailed Behavior of H$\alpha$: Model and Implications}\phantom{+}
\label{subsec:halpha}

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/sf6.pdf}
  \caption{Time-variable fit to H$\alpha$ line profiles.
  {\it Left column:} Raw spectrum at each epoch $f_{\lambda}$ minus
  the time-averaged spectrum $f_{\langle t \rangle}$ (as in
  Figure~\ref{fig:spec}e).
  Underplotted sinusoids are to guide the eye.
  {\it Middle columns:} Model of emission from the inner clump (sum of
  two gaussians) and the outer clump (single gaussian), plotted over
  the data.
  {\it Right column:} Residual of the left column after subtracting
  the sum of the two middle columns, leaving variability in the line
  core.  
  Section~\ref{subsec:halpha} discusses the use of this model.
  }
  \label{fig:halphamodel}
\end{figure}

{\it A model for the time-dynamic H$\alpha$ spectrum}---While
Figure~\ref{fig:spec} shows that clumps of circumstellar material
exist around TIC~141146667, it would be valuable to quantify the exact
orbital periods, velocities, and velocity dispersions of these clumps.
These quantities can clarify the precision with which the
spectroscopic period agrees with the photometric period, and also
constrain the physical dimensions of the emitting region.

Given that a full radiative transfer simulation was beyond our scope,
we opted to construct a phenomenological model aimed at capturing 
the emission from the circumstellar material.  We did this by
fitting each spectral epoch with a multi-component
gaussian, after having subtracted the time-average line profile as in
Figure~\ref{fig:spec}e.  The results of this exercise are shown in
Figure~\ref{fig:halphamodel}; details in its implementation and
interpretation follow.

To the implement the model, we assumed that the ``inner'' ($K_{\rm
inner}$$\approx$2.5\,$v_{\rm eq}$) clump would be well-fit by a sum of
two gaussians because it visually appears double-peaked in the raw
data from $\phi$=0.15-0.35 and $\phi$=0.65-0.85
(Figure~\ref{fig:spec}b).  We assumed the ``outer'' ($K_{\rm
outer}$$\approx$3.9\,$v_{\rm eq}$) clump would be better fit by a
single gaussian, based on its behavior from $\phi$=0.6-0.9.  Each
gaussian component has three free parameters at each spectral epoch:
the mean $\mu$, standard deviation $\sigma$, and amplitude $A$.
Denote the two gaussians for the inner component $i=\{ 0, 1 \}$,
and the single outer component with $i=2$.  Given the complexity of the
line profile data (Figure~\ref{fig:halphamodel} left column), the
likelihood function for this model is multimodal.  We therefore
imposed the prior constraints that $A_i \sim \mathcal{U}[0,1]$,
$\sigma_i/v_{\rm eq} \sim \mathcal{U}[0,1]$, and further assumed
$K_i(t) \sim
\mathcal{U}[
  K_{\rm inner}\sin(\phi(t)) - v_{\rm eq},
  K_{\rm inner}\sin(\phi(t)) + v_{\rm eq}
]$
for the inner two components, and
$K_i(t) \sim
\mathcal{U}[
  K_{\rm outer}\sin(\phi(t) + \pi) - 2v_{\rm eq},
  K_{\rm outer}\sin(\phi(t) + \pi) + 2v_{\rm eq}
]$
for the outer component.
This latter assumption constrains the mean velocity of each component
to be near the value suggested by visual inspection, but provides
sufficient flexibility for the model to deviate from the assumption of
a sinusoid if the data dictate such behavior.
We fitted each component to the data {\it independently} using scipy's
non-linear least squares \texttt{curve\_fit} method
\cite{Virtanen2020}, and scaled the resulting parameter covariance
matrix by a constant factor to match the sample variance of the
residuals.
The choice to fit the $i=\{ 0, 1 \}$, and $i=2$ components
independently generally did not affect the results compared to if we
have fit all three simultaneously, except for at $\phi=1.00$ (see
Figure~\ref{fig:halphamodel}).

The resulting mean velocities, widths, and amplitudes are given in
Table~\ref{tab:halphamodelparams}.
%FIXME TODO gotta get this table being better... and in particular
%gotta not quite signal where there is nothing.  (mask out transits +
%secondary eclipses....)

The main benefit of this model is that it is
sufficiently flexible to capture the observed variability in
circumstellar emission, and

quantify the exact
orbital periods, velocities, and velocity dispersions of these clumps.

Figure~\ref{fig:halphamodel}

functions

{\it Constraints on physical dimensions of emitting region}---

IMPLICATIONS:
The secondary eclipse of the `inner clump' at $\phi$$\approx$0.5 is not
associated with a broadband flux dip.   The clump does not therefore
contribute significant broadband emissivity across the entire TESS
band; only in H$\alpha$.

@@

\begin{sidewaystable}
%\begin{table}
  %\small
  \footnotesize
  \centering
  \begin{tabular}{lccccccccc}
  %\hline 
  %\hline 
  %Time [BJD$_{\rm TDB}$] & RV (\kms) & $\sigma_{\rm RV}$ (\kms) \\
  % from
  % results/halpha_to_rv_timerseries/multigauss_parametertable.tex
  \hline
   & Mean1 & Mean2 & Mean3 & Sigma1 & Sigma2 & Sigma3 & Amplitude1 & Amplitude2 & Amplitude3 \\
  BTJD &  &  &  &  &  &  &  &  &  \\
  \hline
  3357.960099 & -0.725 $\pm$ 0.028 & -1.416 $\pm$ 0.026 & 0.177 $\pm$ 0.019 & 0.39 $\pm$ 0.03 & 0.18 $\pm$ 0.03 & 0.59 $\pm$ 0.02 & 0.51 $\pm$ 0.01 & 0.23 $\pm$ 0.04 & 0.29 $\pm$ 0.01 \\
  3357.970746 & -0.192 $\pm$ 0.007 & 0.498 $\pm$ 0.010 & 0.100 $\pm$ 0.014 & 0.23 $\pm$ 0.01 & 0.24 $\pm$ 0.01 & 0.55 $\pm$ 0.01 & 0.50 $\pm$ 0.01 & 0.37 $\pm$ 0.01 & 0.42 $\pm$ 0.01 \\
  3357.981672 & 0.895 $\pm$ 0.051 & 1.771 $\pm$ 0.076 & -1.470 $\pm$ 0.113 & 0.40 $\pm$ 0.02 & 0.40 $\pm$ 0.10 & 0.09 $\pm$ 0.11 & 0.67 $\pm$ 0.05 & 0.34 $\pm$ 0.04 & 0.03 $\pm$ 0.03 \\
  3357.993524 & 1.536 $\pm$ 0.011 & 2.202 $\pm$ 0.015 & -3.787 $\pm$ 0.049 & 0.25 $\pm$ 0.01 & 0.29 $\pm$ 0.01 & 0.72 $\pm$ 0.05 & 0.55 $\pm$ 0.01 & 0.46 $\pm$ 0.01 & 0.05 $\pm$ 0.00 \\
  3358.003594 & 1.974 $\pm$ 0.007 & 2.588 $\pm$ 0.006 & -2.749 $\pm$ 0.020 & 0.26 $\pm$ 0.01 & 0.22 $\pm$ 0.00 & 1.00 $\pm$ 0.02 & 0.64 $\pm$ 0.01 & 0.67 $\pm$ 0.01 & 0.43 $\pm$ 0.01 \\
  3358.015249 & 2.027 $\pm$ 0.003 & 2.666 $\pm$ 0.003 & -2.370 $\pm$ 0.015 & 0.24 $\pm$ 0.00 & 0.20 $\pm$ 0.00 & 1.00 $\pm$ 0.01 & 0.62 $\pm$ 0.00 & 0.60 $\pm$ 0.00 & 0.56 $\pm$ 0.00 \\
  3358.024971 & 1.790 $\pm$ 0.004 & 2.469 $\pm$ 0.006 & -2.138 $\pm$ 0.011 & 0.24 $\pm$ 0.00 & 0.26 $\pm$ 0.00 & 0.77 $\pm$ 0.01 & 0.53 $\pm$ 0.00 & 0.38 $\pm$ 0.00 & 0.31 $\pm$ 0.00 \\
  3358.036522 & 0.584 $\pm$ 0.008 & 1.399 $\pm$ 0.009 & -1.584 $\pm$ 0.009 & 0.18 $\pm$ 0.01 & 0.40 $\pm$ 0.01 & 0.89 $\pm$ 0.01 & 0.35 $\pm$ 0.01 & 0.39 $\pm$ 0.00 & 0.62 $\pm$ 0.00 \\
  3358.047668 & 0.328 $\pm$ 0.014 & 0.685 $\pm$ 0.010 & -0.969 $\pm$ 0.017 & 0.07 $\pm$ 0.01 & 0.13 $\pm$ 0.01 & 0.46 $\pm$ 0.02 & 0.14 $\pm$ 0.02 & 0.28 $\pm$ 0.02 & 0.52 $\pm$ 0.02 \\
  3358.058824 & -0.646 $\pm$ 0.005 & -1.474 $\pm$ 0.066 & 0.891 $\pm$ 0.079 & 0.23 $\pm$ 0.00 & 0.25 $\pm$ 0.07 & 0.14 $\pm$ 0.08 & 0.51 $\pm$ 0.01 & 0.10 $\pm$ 0.01 & 0.09 $\pm$ 0.04 \\
  3358.069519 & -0.388 $\pm$ 0.104 & -1.669 $\pm$ 0.023 & 2.646 $\pm$ 0.020 & 0.40 $\pm$ 0.07 & 0.40 $\pm$ 0.02 & 0.53 $\pm$ 0.02 & 0.47 $\pm$ 0.04 & 0.33 $\pm$ 0.01 & 0.22 $\pm$ 0.01 \\
  3358.080468 & -1.748 $\pm$ 0.009 & -2.483 $\pm$ 0.010 & 3.526 $\pm$ 0.009 & 0.25 $\pm$ 0.01 & 0.33 $\pm$ 0.01 & 0.55 $\pm$ 0.01 & 0.28 $\pm$ 0.01 & 0.32 $\pm$ 0.00 & 0.32 $\pm$ 0.00 \\
  3358.091255 & -2.119 $\pm$ 0.007 & -2.868 $\pm$ 0.007 & 3.960 $\pm$ 0.005 & 0.26 $\pm$ 0.01 & 0.23 $\pm$ 0.01 & 0.40 $\pm$ 0.01 & 0.34 $\pm$ 0.01 & 0.29 $\pm$ 0.01 & 0.39 $\pm$ 0.00 \\
  3358.101671 & -2.019 $\pm$ 0.006 & -2.731 $\pm$ 0.008 & 3.758 $\pm$ 0.005 & 0.20 $\pm$ 0.01 & 0.24 $\pm$ 0.01 & 0.43 $\pm$ 0.01 & 0.29 $\pm$ 0.01 & 0.25 $\pm$ 0.01 & 0.23 $\pm$ 0.00 \\
  3358.112713 & -1.581 $\pm$ 0.020 & -2.141 $\pm$ 0.043 & 2.867 $\pm$ 0.019 & 0.19 $\pm$ 0.02 & 0.29 $\pm$ 0.03 & 0.40 $\pm$ 0.02 & 0.18 $\pm$ 0.02 & 0.13 $\pm$ 0.01 & 0.12 $\pm$ 0.00 \\
  3358.123928 & -0.624 $\pm$ 0.006 & -1.312 $\pm$ 0.076 & 0.824 $\pm$ 0.012 & 0.09 $\pm$ 0.01 & 0.22 $\pm$ 0.08 & 0.44 $\pm$ 0.01 & 0.18 $\pm$ 0.01 & 0.02 $\pm$ 0.01 & 0.32 $\pm$ 0.01 \\
  3358.135236 & -0.012 $\pm$ 0.011 & -0.208 $\pm$ 0.012 & -0.192 $\pm$ 0.012 & 0.03 $\pm$ 0.01 & 0.11 $\pm$ 0.01 & 0.12 $\pm$ 0.01 & 0.12 $\pm$ 0.04 & 0.25 $\pm$ 0.02 & 0.25 $\pm$ 0.02 \\
  3358.146555 & 1.007 $\pm$ 0.100 & 1.622 $\pm$ 0.048 & -1.789 $\pm$ 0.045 & 0.06 $\pm$ 0.11 & 0.31 $\pm$ 0.06 & 0.04 $\pm$ 0.05 & 0.02 $\pm$ 0.04 & 0.12 $\pm$ 0.02 & 0.05 $\pm$ 0.05 \\
  3358.156693 & 1.603 $\pm$ 0.018 & 2.444 $\pm$ 0.024 & -4.408 $\pm$ 0.197 & 0.14 $\pm$ 0.02 & 0.23 $\pm$ 0.03 & 1.00 $\pm$ 0.17 & 0.14 $\pm$ 0.02 & 0.14 $\pm$ 0.01 & 0.05 $\pm$ 0.00 \\
  3358.169147 & 1.543 $\pm$ 0.007 & 2.152 $\pm$ 0.013 & -3.885 $\pm$ 0.031 & 0.14 $\pm$ 0.01 & 0.39 $\pm$ 0.01 & 1.00 $\pm$ 0.04 & 0.23 $\pm$ 0.01 & 0.27 $\pm$ 0.00 & 0.10 $\pm$ 0.00 \\
  3358.177978 & 1.588 $\pm$ 0.037 & 2.828 $\pm$ 0.005 & -4.378 $\pm$ 0.053 & 0.40 $\pm$ 0.03 & 0.02 $\pm$ 0.01 & 0.75 $\pm$ 0.06 & 0.33 $\pm$ 0.01 & 0.11 $\pm$ 0.03 & 0.12 $\pm$ 0.01 \\
  \hline
  \end{tabular}
  \caption{Multi-gaussian model fit results to H$\alpha$ line profiles.}
  \label{tab:halphamodelparams}
%\end{table}
\end{sidewaystable}
